Equalizer 2.0 --\\Convergence of a Parallel Rendering Framework

### 2018-02-14 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@130 db7c848b-7755-4df2-b763-8796797afa52

Sort-first (\fig{fig:loadeq}) and sort-last load balancing are the most obvious
optimizations for these parallel rendering modes. Our load equalizers are fully
transparent for application developers; that is, they use a reactive approach
based on past rendering times. This assumes a reasonable frame-to-frame
coherency \FIXME{Shouldn't that be frame-to-frame {\em coherence}?}. Equalizer implements two different algorithms, a
\textsf{load\_equalizer} and a \textsf{tree\_equalizer}. The result section
provides some evidence on the strengths and weaknesses of both algorithms.

@@

Compression for the compositing step is critical for performance. This not only
applies to the well-researched network transfer step, but also for the transfer
between GPU and CPU. \textsf{Equalizer} supports a variety of compression
algorithms, from very fast run-length encoding and YUV subsampling on the GPU to JPEG
compression. These algorithms are implemented as runtime-loaded plugins,
allowing easy extension and customization to application-specific compression.
In \cite{MEP:10}, we show this to be a critical step for interactive performance
at scale.

@@

Qt is a popular window system with application developers. Unfortunately, it
imposes a different threading model for window creation and event handling
compared to \textsf{Equalizer}. In \textsf{Equalizer}, each GPU rendering thread
is independently responsible for creating its windows, receiving the events and
eventually dispatching them to the application process' main thread. This design
is motivated by the natural threading model of X11 and WGL, and allows simple
sequential semantics between OpenGL rendering and event handling. In contrast,
Qt requires all windows and QOpenGLContext to be created from the Qt main
thread. An existing Qt window or context may subsequently be moved to a
different thread, and events are signalled from the main thread. For Qt windows,
Equalizer will internally dispatch and handle the window creation from the
render to the main thread, move the created objects back to the render thread,
and dispatch Qt signals to the correct render threads.

@@

\textsf{Sequel} is a simplification layer for \textsf{Equalizer}. It is based on
the realization that while fully expressive, the verbatim abstraction layer of
nodes, pipes, windows and channels in \textsf{Equalizer} requires significant
learning to fully understand and exploit. In reality, a higher abstraction of
only \textsf{Application} and \textsf{Renderer} is sufficient for many use
cases. In \textsf{Sequel}, the application class drives the configuration, and
one renderer instance is created for each (pipe) render thread. They also
provide the natural place to store and distribute data. Last, but not least,
\textsf{ViewData} provides a convenient way to manage multiple views by storing
the camera, model or any other view-specific information.

@@

Distributed objects in \textsf{Collage} provide powerful, object-oriented data
distribution for C++ objects. They facilitate the implementation of data
distribution in a cluster environment. Distributed objects are created by
subclassing from \textsf{co::Serializable} or \textsf{co::Object}. The
application programmer implements serialization and deserialization. Distributed
objects can be static (immutable) or dynamic. Objects have an universally unique
identifier (UUID) as cluster-wide addresses. A master-slave model is used to
establish mapping and data synchronization across processes. Typically, the
application main loop registers a master instance and communicates the UUID to
the render clients, which map their instance to the given identifier. The
following object types are available:

\begin{compactdesc}
\item[Static] The object is not versioned nor buffered. The instance data is
  serialized whenever a new slave instance is mapped. No additional data is
  stored.
\item[Instance] The object is versioned and buffered. The instance and delta
  data are identical; that is, only instance data is serialized. Previous
  instance data is saved to be able to map old versions.
\item[Delta] The object is versioned and buffered. The delta data is typically
  smaller than the instance data. The delta data is transmitted to slave
  instances for synchronization. Previous instance and delta data is saved to be
  able to map and sync old versions.
\item[Unbuffered] The object is versioned and unbuffered. No data is stored, and
  no previous versions can be mapped. No old versions can be mapped and no
  additional data is stored.
\end{compactdesc}

Serialization is facilitated using output or input streams, which abstract the
data transmission and are used like a \textsf{std::stream}. The data streams
implement efficient buffering and compression, and automatically select the best
connection for data transport. Custom data type serializers can be implemented
by providing the appropriate serialization functions. No pointers should be
directly transmitted through the data streams. For pointers, the corresponding
object is typically a distributed object as well, and its UUID and version are
transmitted in place of a pointer.

Dynamic objects are versioned, and on \textsf{commit} the delta data from the
previous version is sent, if available using multicast, to all mapped slave
instances. The data is queued on the remote node, and is applied when the
application calls \textsf{sync} to synchronize the object to a new version. The
\textsf{sync} method might block if a version has not yet been committed or is
still in transmission. All versioned objects the following characteristics:

\begin{compactitem}
\item The master instance of the object generates new versions for all
  slaves. These versions are continuous. It is possible to commit on slave
  instances, but special care has to be taken to handle possible
  conflicts.
\item Slave instance versions can only be advanced; that is, \textsf{sync(
  version )} with a version smaller than the current version will fail.
\item Newly mapped slave instances are mapped to the oldest available
  version by default, or to the version specified when calling
  \textsf{mapObject}.
\end{compactitem}

@@

The most obvious one is compression. Recently, many new compression algorithms
have been developed which exploit modern CPU architectures and deliver
compression rates well above one gigabyte per second. \textsf{Collage} uses the
Pression library~\cite{pression}, which provides an unified interface for a
number of compression libraries, such as FastLZ~\cite{jesperfast},
Snappy~\cite{snappy} and ZStandard~\cite{zstd}. It also contains a custom,
virtually zero-cost RLE compressor. Pression parallelizes the compression and
decompression using data decomposition. This compression is generic, and
implemented transparently for the application. Applications can also use
data-specific compression.

@@

Caching retains instance data of objects in a client-side cache, and reuses this
data to accelerate mapping of objects. The instance cache is either filled by
``snooping'' on multicast transmissions or by an explicit preloading when the
master objects are registered. The preloading sends instance data of recently
registered master objects while the corresponding node is idle to all connected
nodes. These nodes simply enter the received data to their cache. Preloading
uses multicast when available.

@@

This figure gives an impression of how individual modes perform on heterogeneous
systems. In this case the tree equalizer performs best (\fig{fEqualizers}
(left)), as it allows us to a priori define how much {\em usage} it should make
of individual nodes, i.e., bias the allocation of rendering time in accordance
with the (simulated) compute resources. \fig{fEqualizers} (right), on the other
hand, illustrates how the tested decomposition modes perform on a system where
compute resources fluctuate randomly every frame, as can arguably be the case
for virtualized enviroments. For this scenario tile equalizer seems best suited,
as it performs load balancing implicitly and does not assume coherence of
available resources between frames. The simpler tree equalizer also outperforms
the load equalizer in this experiment.

@@

A critical piece for data distribution performance are the characteristics of
the data compression algorithm. Our microbenchmark compresses a set of binary
files, precalculating the speed and compression ratio of the various engines.
\fig{fCompressor} shows the compression and decompression speed in gigabyte per
second as well as the size of the compressed data relative to the uncompressed
data. The ZSTD$x$ engines use the ZStandard compression library at compression
level $x$. The measurements were performed on a single, isolated node.

@@

In this section we analyse how data distribution and synchronization performs in
real-world applications. We extracted the existing data distribution code from a
mesh renderer (eqPly) and a volume renderer (Livre) into a benchmark application
to measure the time to initially map all the objects on the render client nodes,
and to perform a commit-sync of the full data set after mapping has been
established. All figures observe a noticable measurement jitter due to other
services running on the cluster. The details of the benchmark algorithm can be
found in the
\htmladdnormallinkfoot{implementation}{https://github.com/Eyescale/Equalizer/tree/paper2018/tools/eqObjectBench}.
used the same data sets as in the previous section. We ran the benchmark on up
to eight physical nodes, that is, after eight processes nodes start to run two
processes per node, which share CPU, memory and network interface bandwidth.

>>>>

Sort-first (\fig{fig:loadeq}) and sort-last load balancing are the most obvious
optimizations for these parallel rendering modes. Our load equalizers are fully
transparent for application developers; that is, they use a reactive approach
based on past rendering times. This assumes a reasonable frame-to-frame
coherency. \FIXME{Shouldn't that be frame-to-frame {\em coherence}?} Equalizer implements two different algorithms, a
\textsf{load\_equalizer} and a \textsf{tree\_equalizer}. The result section
provides some evidence on the strengths and weaknesses of both algorithms.

@@

Compression for the compositing step is critical for performance. This not only
applies to the well-researched network transfer step, but also for the transfer
between GPU and CPU. \textsf{Equalizer} supports a variety of compression
algorithms, from very fast run-length encoding (RLE) and YUV subsampling on the GPU to JPEG
compression. These algorithms are implemented as runtime-loaded plugins,
allowing easy extension and customization to application-specific compression.
In \cite{MEP:10}, we show this to be a critical step for interactive performance
at scale.

@@

Qt is a popular window system with application developers. Unfortunately, it
imposes a different threading model for window creation and event handling
compared to \textsf{Equalizer}. In \textsf{Equalizer}, each GPU rendering thread
is independently responsible for creating its windows, receiving the events and
eventually dispatching them to the application process' main thread. This design
is motivated by the natural threading model of X11 and WGL, and allows simple
sequential semantics between OpenGL rendering and event handling. In contrast,
Qt requires all windows and each QOpenGLContext to be created from the Qt main
thread. An existing Qt window or context may subsequently be moved to a
different thread, and events are signalled from the main thread. For Qt windows,
Equalizer will internally dispatch and handle the window creation from the
render to the main thread, move the created objects back to the render thread,
and dispatch Qt signals to the correct render threads.

@@

\textsf{Sequel} is a simplification layer for \textsf{Equalizer}. It is based on
the realization that while fully expressive, the verbatim abstraction layer of
nodes, pipes, windows and channels in \textsf{Equalizer} requires significant
learning to fully understand and exploit. In reality, a higher abstraction of
only \textsf{Application} and \textsf{Renderer} is sufficient for many use
cases. In \textsf{Sequel}, the application class drives the configuration, and
one renderer instance is created for each (pipe) render thread. They also
provide the natural place to store and distribute data. Finally,
\textsf{ViewData} provides a convenient way to manage multiple views by storing
the camera, model or any other view-specific information.

@@

Distributed objects in \textsf{Collage} provide powerful, object-oriented data
distribution for C++ objects. They facilitate the implementation of data
distribution in a cluster environment. Distributed objects are created by
subclassing from \textsf{co::Serializable} or \textsf{co::Object}. The
application programmer implements serialization and deserialization. Distributed
objects can be static (immutable) or dynamic. Objects have a universally unique
identifier (UUID) as cluster-wide address. A master-slave model is used to
establish mapping and data synchronization across processes. Typically, the
application main loop registers a master instance and communicates the UUID to
the render clients, which map their instance to the given identifier. The
following object types are available:

\begin{compactdesc}
\item[Static] The object is not versioned nor buffered. The instance data is
  serialized whenever a new slave instance is mapped. No additional data is
  stored.
\item[Instance] The object is versioned and buffered. The instance and delta
  data are identical; that is, only instance data is serialized. Previous
  instance data is saved to be able to map old versions.
\item[Delta] The object is versioned and buffered. The delta data is typically
  smaller than the instance data. The delta data is transmitted to slave
  instances for synchronization. Previous instance and delta data is saved to be
  able to map and sync old versions.
\item[Unbuffered] The object is versioned and unbuffered. No data is stored, and
  no previous versions can be mapped. \FIXME{Redundant.} No old versions can be mapped and no
  additional data is stored.
\end{compactdesc}

Serialization is facilitated using output or input streams, which abstract the
data transmission and are used like an \textsf{std::stream}. The data streams
implement efficient buffering and compression, and automatically select the best
connection for data transport. Custom data type serializers can be implemented
by providing the appropriate serialization functions. No pointers should be
directly transmitted through the data streams. For pointers, the corresponding
object is typically a distributed object as well, and its UUID and version are
transmitted in place of a pointer.

Dynamic objects are versioned, and on \textsf{commit} the delta data from the
previous version is sent, if available using multicast, to all mapped slave
instances. The data is queued on the remote node, and is applied when the
application calls \textsf{sync} to synchronize the object to a new version. The
\textsf{sync} method might block if a version has not yet been committed or is
still in transmission. All versioned objects have the following characteristics:

\begin{compactitem}
\item The master instance of the object generates new versions for all
  slaves. These versions are continuous. It is possible to commit on slave
  instances, but special care has to be taken to handle possible
  conflicts.
\item Slave instance versions can only be advanced; that is, \textsf{sync(
  version)} with a version smaller than the current version will fail.
\item Newly mapped slave instances are mapped to the oldest available
  version by default \FIXME{Is this correct?}, or to the version specified when calling
  \textsf{mapObject}.
\end{compactitem}

@@

The most obvious one is compression. Recently, many new compression algorithms
have been developed which exploit modern CPU architectures and deliver
compression rates well above one gigabyte per second. \textsf{Collage} uses the
Pression library~\cite{pression}, which provides a unified interface for a
number of compression libraries, such as FastLZ~\cite{jesperfast},
Snappy~\cite{snappy} and ZStandard~\cite{zstd}. It also contains a custom,
virtually zero-cost RLE compressor. Pression parallelizes the compression and
decompression using data decomposition. This compression is generic, and
implemented transparently for the application. Applications can also use
data-specific compression.

@@

Caching retains instance data of objects in a client-side cache, and reuses this
data to accelerate mapping of objects. The instance cache is either filled by
``snooping'' on multicast transmissions or by an explicit preloading when the
master objects are registered. The preloading sends instance data of recently
registered master objects to all connected
nodes, while the corresponding node is idle. These nodes simply enter the received data to their cache. Preloading
uses multicast when available.

@@

This figure gives an impression of how individual modes perform on heterogeneous
systems. In this case the tree equalizer performs best (\fig{fEqualizers}
(left)), as it allows us to a priori define how much {\em usage} it should make
of individual nodes, i.e., bias the allocation of rendering time in accordance
with the (simulated) compute resources. \fig{fEqualizers} (right), on the other
hand, illustrates how the tested decomposition modes perform on a system where
compute resources fluctuate randomly every frame, as can arguably be the case
for shared rendering nodes in virtualized enviroments. For this scenario tile equalizer seems best suited,
as it performs load balancing implicitly and does not assume coherence of
available resources between frames. The simpler tree equalizer also outperforms
the load equalizer in this experiment.

@@

A critical piece for data distribution performance are the characteristics of
the data compression algorithm. Our microbenchmark compresses a set of binary
files \FIXME{What are these files? Fig. 7 only speaks of "generic" binary data...}, precalculating the speed and compression ratio of the various engines.
\fig{fCompressor} shows the compression and decompression speed in gigabyte per
second as well as the size of the compressed data relative to the uncompressed
data. The ZSTD$x$ engines use the ZStandard compression library at compression
level $x$. The measurements were performed on a single, isolated node.

@@

In this section we analyse how data distribution and synchronization performs in
real-world applications. We extracted the existing data distribution code from a
mesh renderer (eqPly) and a volume renderer (Livre) into a benchmark application
to measure the time to initially map all the objects on the render client nodes,
and to perform a commit-sync of the full data set after mapping has been
established. All figures observe a noticable measurement jitter due to other
services running on the cluster. The details of the benchmark algorithm can be
found in the
\htmladdnormallinkfoot{implementation}{https://github.com/Eyescale/Equalizer/tree/paper2018/tools/eqObjectBench}.

\FIXME{Sentence incomplete.}
used the same data sets as in the previous section. We ran the benchmark on up
to eight physical nodes, that is, after eight processes nodes start to run two
processes per node, which share CPU, memory and network interface bandwidth.
